---
title: "DS C5 W2 Project 2"
author: "Ksenia Kharitonova"
date: "21 April 2017"
output: 
  html_document: 
    keep_md: yes
---

## Loading and preprocessing the data

Loading necessary packages.

```{r packages, message=FALSE}
library(dplyr)
library(ggplot2)
```

Loading data. The file has to be present in the same directory as RMD file.

```{r data}
activity <- read.csv("activity.csv")
```

Looking at the description and the summary.

```{r summary}
str(activity)
summary(activity)
```

We can see that variable `steps` has `2304` missing values. We will see what we can do with it. The `date` variable is represented by a factor with levels as strings. Let's transform it into a date variable.

```{r date transform}
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

Calculating the total number of steps taken per day. For the moment we ignore missing values.

```{r total-steps}
total_steps <- activity %>% group_by(date) %>% summarise(steps_t = sum(steps, na.rm = TRUE))
```

Histogram of total number of steps per day.

```{r histogram-total-steps}
ggplot(total_steps, aes(steps_t)) + geom_histogram(bins = 30)+labs(x = "Total Steps", y = "Count")
```

The median and the mean of total steps per day.

```{r mean-median-total-steps}
total_steps %>% summarise(mean = mean(steps_t), median = median(steps_t))
```

## What is the average daily activity pattern?

Calculating the average number of steps at the interval across all days.

```{r average_steps}
avg_steps <- activity %>% group_by(interval) %>% summarise(avg_steps = mean(steps, na.rm=TRUE))
```

Time series line plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

```{r plot-avg-steps}
ggplot(avg_steps, aes(interval, avg_steps)) + geom_line()+labs(x = "Interval", y = "Average steps")
```

5-minute interval, on average across all the days in the dataset, with the maximum number of steps.

```{r interval-max-avg-steps}
avg_steps %>% summarise(interval = interval[which.max(avg_steps)], max = max(avg_steps))
```

## Imputing missing values

Calculate and report the total number of missing values in the dataset.

```{r number-na}
summary(activity)
```

We can see that there is `2304` missing values in the variable `steps`.

Let's replace the missing values by the median of the steps on that interval across all days in a new dataset `activity_new`.

```{r replacing-na}
activity_new <- activity %>% group_by(interval) %>% mutate(steps = ifelse(is.na(steps), median(steps, na.rm = TRUE), steps))
```

Let's calculate total steps with the new dataset.

```{r total-steps-new}
total_steps_new <- activity_new %>% group_by(date) %>% summarise(steps_t = sum(steps))
```

Histogram of total number of steps per day on a new dataset.

```{r histogram-total-steps-new}
ggplot(total_steps_new, aes(steps_t)) + geom_histogram(bins = 30)+labs(x = "Total Steps", y = "Count")
```

The median and the mean of total steps per day on a new dataset.

```{r}
total_steps_new %>% summarise(mean = mean(steps_t), median = median(steps_t))
```

We can see that our strategy did not change the median od the total steps but the mean has decreased.

## Are there differences in activity patterns between weekdays and weekends?

Creating a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r weekdays-weekends}
activity_new <- activity_new %>% mutate(wd = factor(ifelse(weekdays(date, abbreviate = FALSE) %in% c("Saturday","Sunday"),1,0), labels = c("Weekday", "Weekend")))
```

Calculating the average number of steps at the intervals across weekdays and weekends.

```{r average-steps}
avg_steps_new <- activity_new %>% group_by(wd, interval) %>% summarise(avg_steps = mean(steps, na.rm=TRUE))
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
ggplot(avg_steps_new, aes(interval, avg_steps)) + facet_wrap( ~ wd , ncol=1) + geom_line()+labs(x = "Interval", y = "Average steps")
```
